Trends in Mortality Among Shelter Residents: : Closely Tied To Virus Outbreaks

### 2024-09-27 paper/paper.qmd
change title and intro to better fit the rubric

---
title: "Trends in Mortality Rates"
subtitle: "Analyzing Deaths Among Shelter Residents from 2007 to 2024"
author: 
  - Mandy He
thanks: "Code and data are available at: https://open.toronto.ca/dataset/deaths-of-shelter-residents/"
date: Septmeber 23 2024
date-format: long
abstract: "Poverty and homelessness remain pressing issues in today’s society, particularly evident in downtown Toronto, where many individuals still sleep on the streets or in subway stations. Due to these harsh living conditions, these individuals often face severe health issues, leading to an increased risk of mortality. Luckily, there are services in the city that provide place for some of these people to live. This paper analyzes the trends in mortality rates among shelter residents in the City of Toronto from 2007 to 2024, highlighting patterns over time and exploring the factors contributing to these rates."
format: pdf
number-sections: true
bibliography: references.bib 
toc: true

@@

library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(tidyr)
library(gridExtra)

Death_data <- read_csv(here::here("data/analysis_data/Num_death_analysis_data.csv"))
AvgAge_data <- read_csv(here::here("data/analysis_data/AvgAge_death_analysis_data.csv"))

@@

The data was examined more in depth by analyzing the mortality rate and average mortality rate between male and female homeless individuals in shelter residences. There are a lot more male death compare to female death every single year. Additionally, the average age at death is around 50 to 60 years old, which is relatively young compared to Canada’s life expectancy. Future details about the findings will be discussed in the discussion section.

@@

The data used in this paper was collected by the Toronto Shelter and Support Services Division since 2007. Shelters are required to notify the City of Toronto within 24 hours of a death and submit a written report within 30 days. Consequently, the data is updated monthly on the Open Data Toronto website. The raw data set includes 10 variables and 18 observations, covering the period from 2007 to 2024, as shown in @tbl-Table1. The key variables used in this paper include: year, total decedents, average age at death for all decedents, male decedents, average age at death for male decedents, female decedents, and average age at death for female decedents. Additionally, the monthly death counts from 2007 to 2024 will be used for graphing, comprising 212 observations.

@@

\newpage
```{r}
#| label: tbl-Table1
#| tbl-cap: Trends in Mortality Rates Among Shelter Residents (2007-2024)
#| echo: false

@@

AvgAge_data |> 
  ggplot(aes(x = year, y = total_decedents)) +
  geom_point() +
  geom_line() +
  labs(title = "Total Shelter Residents Decedents from 2007 to 2024", x = "Year", y = "Total Decedents") +
  theme_minimal() +
   theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5)) + 
  scale_x_continuous(breaks = unique(AvgAge_data$year)) +
  scale_y_continuous(breaks = seq(0, max(AvgAge_data$total_decedents), by = 20))

@@

\newpage
@fig-TotalDecedentbyMonthcovid provides a close-up view of the years 2019 to 2024, covering the period of the COVID-19 pandemic and a bit of the post-pandemic phase. During this time, the number of deaths does not exhibit a clear trend as time progresses. However, in many months during these years, the number of deaths remains notably high, often fluctuating between 10 and 15 deaths per month. Detailed discussion found in @sec-second-dpoint.
```{r}
#| label: fig-TotalDecedentbyMonthcovid
#| fig-cap: Yearly Mortality rate from 2019-2023
#| echo: false

# Filter the data for years 2019 to 2022
Death_data_filtered <- Death_data %>%
  filter(year >= 2019 & year <= 2023)

@@

# Plot the graph for filtered data
Death_data_filtered |>
  ggplot(aes(x = YearMonth, y = total_decedents, group = 1)) +
  geom_line(color = "black") +
  geom_point(color = "black") +
  scale_x_discrete(breaks = Death_data_filtered$YearMonth[seq(1, nrow(Death_data_filtered), by = 6)]) + # Show every 6th month
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) + # Display x-axis labels vertically
  labs(title = "Trends in Total Decedents by Month(2019-2023)",
       x = "Month-Year",
       y = "Total Decedents") +
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5))

```
\newpage
@fig-T illustrates the number of deaths among genders. The bar graph clearly shows that the number of male decedents is significantly greater than that of female decedents, a trend consistent across all the years presented. Additionally, the graph indicates that the mortality rate among males is more volatile over the years, particularly from 2019 to 2022. These results may be attributed to the proportion of males living in shelters; further details can be found in @sec-3-dpoint.
```{r}
#| label: fig-T 
#| fig-cap: Total Decedents Female VS Male
#| echo: false
#| warning: false
#| message: false

# Reshape the data for plotting
plot_data <- AvgAge_data %>%
  select(year, Male_decedents, Female_decedents) %>%
  pivot_longer(cols = c(Male_decedents, Female_decedents),
               names_to = "Gender",
               values_to = "Mortality")

# Create a new variable for the last two digits of the year with leading zero
plot_data <- plot_data %>%
  mutate(year_short = sprintf("%02d", year %% 100))

# Create the bar graph with adjusted year labels
ggplot(plot_data, aes(x = year_short, y = Mortality, fill = Gender)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  labs(title = "Mortality Numbers of Male Vs Female Decedents over 2007 to 2024",
       x = "Year (2000s)",
       y = "Number of Decedents") +
  scale_fill_manual(values = c("Male_decedents" = "blue", "Female_decedents" = "pink")) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  theme(plot.title = element_text(hjust = 0.5))

@@

```{r}
#| label: fig-AvgAgeofDecedentsFemaleVSMale 
#| fig-cap: Total Decedents Female VS Male
#| echo: false
#| warning: false
#| message: false

# Reshape the data for plotting average age of death
avg_age_data <- AvgAge_data %>%
  select(year, Average_age_of_death_male_decedents, Average_age_of_death_female_decedents) %>%
  pivot_longer(cols = c(Average_age_of_death_male_decedents, Average_age_of_death_female_decedents),
               names_to = "Gender",
               values_to = "Average_Age") %>%
  mutate(year_short = sprintf("%02d", year %% 100))

# Create the line graph for average age of death
ggplot(avg_age_data, aes(x = year_short, y = Average_Age, color = Gender, group = Gender)) +
  geom_line(size = 1) +   
  geom_point(size = 2) +    
  labs(title = "Average Age of Death for Male Vs Female Decedents",
       x = "Year",
       y = "Average Age of Death") +
  scale_color_manual(values = c("Average_age_of_death_male_decedents" = "blue", 
                                 "Average_age_of_death_female_decedents" = "pink")) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  theme(plot.title = element_text(hjust = 0.5))

@@

The findings in this paper represent a small aspect of what is happening in the world, focusing specifically on the homeless population. While these results cannot be used to draw definitive conclusions about broader trends, they do provide insights into patterns that may be occurring on a larger scale.

@@

Death_data$month <- factor(Death_data$month, levels = month.abb)

@@

plot_2011 <- ggplot(current_year_data_2011, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(title = "Total Decedents by Month in 2011", 
       x = "Month", 
       y = "# Decedents") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

>>>>

---
title: "Trends in Mortality Among Shelter Residents: : Closely Tied To Virus Outbreaks"
subtitle: "Analyzing period: 2007 to 2024"
author: 
  - Mandy He
thanks: "Code and data are available at: https://open.toronto.ca/dataset/deaths-of-shelter-residents/"
date: Septmeber 27 2024
date-format: long
abstract: "Homelessness remain pressing issues in today’s society, particularly evident in downtown Toronto, due to  harsh living conditions, these individuals often face severe health issues, leading to an increased risk of mortality. This paper analyzes the trends in mortality rates among shelter residents in the City of Toronto from 2007 to 2024, highlighting patterns over time and exploring the factors contributing to these rates. This paper found that the trend in the mortality rate of this minority group reflects the overall health conditions in the macro-environment. Therefore, examining these patterns can help improve provisions and reduce the number of deaths among both this vulnerable population and society at large."
format: pdf
number-sections: true
bibliography: references.bib 
toc: true

@@

#load necessary package 
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(tidyr)
library(gridExtra)

# load data
Death_data <- read_csv(here::here("data/analysis_data/Num_death_analysis_data.csv"))
AvgAge_data <- read_csv(here::here("data/analysis_data/AvgAge_death_analysis_data.csv"))

@@

The data was examined more in depth by analyzing the mortality rate and average mortality rate between male and female homeless individuals in shelter residences. There are a lot more male death compare to female death every single year. Additionally, the average age at death is around 50 to 60 years old, which is relatively young compared to Canada’s life expectancy. These trends are crucial for helping the city improve services provided to this population. Furthermore, from a social and economic perspective, supporting minority populations is essential for community and economic growth. If individuals in these communities are healthier and live longer, they can contribute more effectively to society.

@@

The data used in this paper was collected by the Toronto Shelter and Support Services Division since 2007. Shelters are required to notify the City of Toronto within 24 hours of a death and submit a written report within 30 days. Accordingly, the data is updated monthly on the Open Data Toronto website. The raw data set includes 10 variables and 18 observations, covering the period from 2007 to 2024, as shown in @tbl-Table1. The key variables used in this paper include: year, total decedents, average age at death for all decedents, male decedents, average age at death for male decedents, female decedents, and average age at death for female decedents. Additionally, the monthly death counts from 2007 to 2024 will be used for graphing, comprising 212 observations.

@@

\newpage
```{r}
#| label: tbl-Table1
#| tbl-cap: Variable In The Data - Detah Among Shelter Residents (2007-2024)
#| echo: false

@@

# creating a line graph 
AvgAge_data |> 
  ggplot(aes(x = year, y = total_decedents)) +
  geom_point() +
  geom_line() +
  labs(title = "Total Shelter Residents Decedents from 2007 to 2024", x = "Year", y = "Total Decedents") +
  theme_minimal() +
   theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5)) + 
  scale_x_continuous(breaks = unique(AvgAge_data$year)) +
  scale_y_continuous(breaks = seq(0, max(AvgAge_data$total_decedents), by = 20))

@@

\newpage
@fig-TotalDecedentbyMonthcovid provides a close-up view of the years 2019 to 2024, covering the period of the COVID-19 pandemic and a bit of the post-pandemic phase. During this time, the number of deaths does not exhibit a clear trend as time progresses. However, in many months during these years, the number of deaths remains notably high, often fluctuating between 10 and 15 deaths per month. Detailed discussion found in @sec-second-dpoint.
```{r}
#| label: fig-TotalDecedentbyMonthcovid
#| fig-cap: Yearly Mortality rate during COVID (2019-2023)
#| echo: false

# Filter the data for years 2019 to 2022
Death_data_filtered <- Death_data |> 
  filter(year >= 2019 & year <= 2023)

@@

# Plot the graph for filtered data
Death_data_filtered |>
  ggplot(aes(x = YearMonth, y = total_decedents, group = 1)) +
  geom_line(color = "black") +
  geom_point(color = "black") +
  scale_x_discrete(breaks = Death_data_filtered$YearMonth[seq(1, nrow(Death_data_filtered), by = 6)]) + 
  labs(title = "Trends in Total Decedents by Month(2019-2023)",
       x = "Month-Year",
       y = "Total Decedents") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) + 
  theme(plot.title = element_text(hjust = 0.5))

```
\newpage
@fig-T illustrates the number of deaths among genders. The bar graph clearly shows that the number of male decedents is significantly greater than that of female decedents, a trend consistent across all the years presented. Additionally, the graph indicates that the mortality rate among males is more volatile over the years, particularly from 2019 to 2022. These results may be attributed to the proportion of males living in shelters; further details can be found in @sec-3-dpoint.
```{r}
#| label: fig-T 
#| fig-cap: Total Decedents Split by Gender (Female VS Male)
#| echo: false
#| warning: false
#| message: false

# Reshape the data for plotting
plot_data <- AvgAge_data |>
  select(year, Male_decedents, Female_decedents)|>
  pivot_longer(cols = c(Male_decedents, Female_decedents),
               names_to = "Gender",
               values_to = "Mortality")

# Create a new variable for the last two digits of the year with leading zero
plot_data <- plot_data |>
  mutate(year_short = sprintf("%02d", year %% 100))

# Create the bar graph with adjusted year labels
plot_data |>
  ggplot(aes(x = year_short, y = Mortality, fill = Gender)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  labs(title = "Mortality Numbers of Male Vs Female Decedents over 2007 to 2024",
       x = "Year (2000s)",
       y = "Number of Decedents") +
  scale_fill_manual(values = c("Male_decedents" = "blue", "Female_decedents" = "pink")) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  theme(plot.title = element_text(hjust = 0.5))

@@

```{r}
#| label: fig-AvgAgeofDecedentsFemaleVSMale 
#| fig-cap: Average Age of Death (Female VS Male)
#| echo: false
#| warning: false
#| message: false

# Reshape the data for plotting average age of death
avg_age_data <- AvgAge_data |>
  select(year, Average_age_of_death_male_decedents, Average_age_of_death_female_decedents) %>%
  pivot_longer(cols = c(Average_age_of_death_male_decedents, Average_age_of_death_female_decedents),
               names_to = "Gender",
               values_to = "Average_Age") |>
  mutate(year_short = sprintf("%02d", year %% 100))

# Create the line graph for average age of death
avg_age_data |>
  ggplot(aes(x = year_short, y = Average_Age, color = Gender, group = Gender)) +
  geom_line(size = 1) +   
  geom_point(size = 2) +    
  labs(title = "Average Age of Death for Male Vs Female Decedents",
       x = "Year",
       y = "Average Age of Death") +
  scale_color_manual(values = c("Average_age_of_death_male_decedents" = "blue", 
                                 "Average_age_of_death_female_decedents" = "pink")) +
  theme_minimal() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  theme(plot.title = element_text(hjust = 0.5))

@@

The findings in this paper represent a small aspect of what is happening in the world, focusing specifically on the homeless population. While these results cannot be used to draw definitive conclusions about broader trends, they do provide awareness into patterns that may be occurring on a larger scale.

@@

# make sure that the month is in order from Jan to Dec.
Death_data$month <- factor(Death_data$month, levels = month.abb)

@@

# Create the bar plots
plot_2011 <- ggplot(current_year_data_2011, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(title = "Total Decedents by Month in 2011", 
       x = "Month", 
       y = "# Decedents") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))
